---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Presence-absence data

In this chapter we first, download occurrence data from global open-access datasets such as Global Biodiversity Information Facility (GBIF, https://www.gbif.org/) and Ocean Biodiversity Information System (OBIS, https://obis.org/); second, clean downloaded data reformating, renaming fields and removing outliers data; and lastly, we generate a set of pseudoabsence points along our study area. 

First we load list required libraries.

```{r, eval=T,message=FALSE,warning=FALSE}
requiredPackages <- c(
  #GENERAL USE LIBRARIES --------#
  "here", # Library for reproducible workflow
  "rstudioapi",  # Library for reproducible workflow
  "maptools", #plotting world map
  "ggplot2", #for plotting
  
  #Presence-absence data chapter--------#
  "robis", # Specific library to get the occurrence data
  "rgbif",# Specific library to get the occurrence data
  "CoordinateCleaner", #to remove outlier
  "rgdal", # to work with Spatial data
  "sf", # to work with spatial data (shapefiles)
  "data.table", #for reading data,
  "dplyr", #for reading data,
  "tidyr" #for reading data
  )
```

Function to install the required packages that are not in your system and load all the required packages
```{r, eval=F}
install_load_function <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

install_load_function(requiredPackages)

```

Define some settings
```{r, eval=F}
# location of script
setwd(dirname(getSourceEditorContext()$path))

# general settings for ggplot (black-white background, larger base_size)
theme_set(theme_bw(base_size = 16))
```

## Download presence data

First we need to define our study area, in this case we select the Atlantic ocean based on the The Food and Agriculture Organization (FAO) Major Fishing Areas for Statistical Purposes and we remove Black sea subarea.   
```{r, eval=F}
# url where FAO shapfile is stored
url<-"https://www.fao.org/fishery/geoserver/area/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=area%3AFAO_AREAS&maxFeatures=50&outputFormat=SHAPE-ZIP"

#download file
download.file(url,"data/spatial/FAO_AREAS.zip",mode="wb")

#unzip downloaded file
unzip("data/spatial/FAO_AREAS.zip",exdir="data/spatial")

# Load FAO (spatial multipolygon)
FAO<- st_read(file.path("data", "spatial", "FAO_AREAS.shp"))

#Select Atlantic Ocean FAO Area 
FAO_Atl <- FAO[FAO$OCEAN=="Atlantic",]

#Select Black sea subarea 
Black_Sea <- FAO_Atl[FAO_Atl$ID=="20",]

# transform to sf objects
FAO_Atl.sf <- st_as_sf(FAO_Atl)
Black_Sea.sf <- st_as_sf(Black_Sea)

# Remove Black sea using st_difference (reverse of st_intersection)
FAO_Atl_no_black_sea <- st_difference(FAO_Atl.sf,Black_Sea.sf) %>%   dplyr::select (F_AREA)

#transform to spatial polygons dataframe
study_area<- sf:::as_Spatial(FAO_Atl_no_black_sea)

plot(study_area)
```

Download data from OBIS and GBIF using scientific name. In this case we select Albacore tuna species (<em>Thunnus alalunga</em>).
```{r, eval=F}
# Get data from OBIS
mydata.obis<-robis::occurrence(scientificname="Thunnus alalunga")

# Get data from GBIF
mydata.gbif<-occ_data(scientificName="Thunnus alalunga", hasCoordinate = TRUE, limit=100000)$data
```

We now check the downloaded data and select the fields of interest
```{r, eval=F}
# check names for GBIF data
names(mydata.gbif)
  
#select columns of interest
mydata.gbif <- mydata.gbif %>%
                dplyr::select("acceptedScientificName",
                  "decimalLongitude",
                  "decimalLatitude",
                  "year",
                  "month",
                  "day",
                  "eventDate",
                  "depth")

#check names in for OBIS data
names(mydata.obis)

#select columns of interest
mydata.obis <-  mydata.obis %>%
                dplyr::select("scientificName",
                  "decimalLongitude",
                  "decimalLatitude",
                  "date_year",
                  "month",
                  "day",
                  "eventDate",
                  "depth",
                  "bathymetry",
                  "occurrenceStatus",
                  "sst")
  
```

Reformat the data addin a new field and renaming some columns from mydata.gbif dataframe in order to have the same columns and be able to join both tables

```{r, eval=F}
mydata.gbif <- mydata.gbif %>% 
    dplyr::rename(scientificName= "acceptedScientificName") %>% 
    dplyr::rename(date_year = "year") %>% 
    dplyr::mutate(bathymetry= NA) %>% 
    dplyr::mutate(occurrenceStatus=1) %>% 
    dplyr::mutate(sst= NA)

#Join data from OBIS and GBIF 
mydata.fus<-rbind(mydata.obis,mydata.gbif)
  
#assign unique scientific name 
  mydata.fus <- mydata.fus %>% 
    dplyr::mutate(scientificName= paste(mydata.obis$scientificName[1]))
```

Clean raw data
```{r, eval=F}
#give date format to eventDate and fill out month and date_year columns
  mydata.fus$eventDate <- as.Date(mydata.fus$eventDate)
  mydata.fus$date_year <- as.numeric(mydata.fus$date_year)
  mydata.fus$month <- as.numeric(mydata.fus$month)

#mutate occurrenceStatus column giving value of 1 to presences and 0 to absences
  
unique (mydata.fus$occurrenceStatus)

mydata.fus <- mydata.fus %>% 
    mutate(occurrenceStatus = ifelse(occurrenceStatus== "NA", NA, occurrenceStatus)) %>%
    mutate(occurrenceStatus = ifelse(occurrenceStatus== "Present", 1, occurrenceStatus)) %>% 
    mutate(occurrenceStatus = ifelse(occurrenceStatus== "present", 1, occurrenceStatus)) %>% 
    mutate(occurrenceStatus = ifelse(occurrenceStatus== "Presente", 1, occurrenceStatus)) %>% 
    mutate(occurrenceStatus = ifelse(occurrenceStatus== "Presence", 1, occurrenceStatus)) %>% 
    mutate(occurrenceStatus = ifelse(occurrenceStatus== "P", 1, occurrenceStatus)) %>% 
    mutate(occurrenceStatus = ifelse(occurrenceStatus== "Q", 1, occurrenceStatus))
```

Assign 1 value to all occurrences
```{r, eval=F}
mydata.fus <- mydata.fus  %>% 
    dplyr::mutate(occurrenceStatus = 1)
```

Remove outliers based on distance method (total distance= 1000 km)
```{r, eval=F}

out.dist <- cc_outl(x=mydata.fus,
                lon = "decimalLongitude", lat = "decimalLatitude",
                species = "scientificName",
                method="distance", tdi=1000, # distance method with tdi=1000km
                thinning=T, thinning_res=0.5,
                value="flagged") 

#remove outliers from the data
mydata.fus <- mydata.fus[out.dist, ]
```

Remove duplicates based on the date
```{r, eval=F}

date <- cbind(mydata.fus$decimalLongitude,mydata.fus$decimalLatitude,mydata.fus$eventDate)

mydata.fus<-mydata.fus[!duplicated(date),]
```

Prepare the data to 
```{r, eval=F}
# PREPARE DATA TO USE FAO ATLANTIC REGION MASK
---------------------------------    
    
  # Prepare coordinate format and projection to be able to use FAO zone masks
  dat <- data.frame(cbind(mydata.fus$decimalLongitude,mydata.fus$decimalLatitude))
  ptos<-as.data.table(dat,keep.columnnames=TRUE)
  
  coordinates(ptos) <- ~ X1 + X2
  
  proj4string(ptos) <-proj4string(study_area)
  
  ## Select only occurrences from FAO Atlantic
  match2<-data.frame(subset(mydata.fus,!is.na(over(ptos, study_area)[,1])))
  
  ## Extract the FAO area of each point
  match3<-data.frame(subset(over(ptos, study_area), !is.na(over(ptos, study_area)[,1])))
  
  # create data frame with area, name, lon, lat and year 
  df0<-cbind(F_AREA=match3$F_AREA,match2)[,c("F_AREA","scientificName","decimalLongitude","decimalLatitude","date_year","occurrenceStatus")]
  
  #rename some columns
  names(df0)[3:5]<-c("LON","LAT","YEAR")
  
  # add bathymetry from NOAA
  library(marmap)

  bathy <- getNOAA.bathy(lon1=-100,lon2=30,lat1=-41,lat2=55, resolution = 1, 
                       keep=FALSE, antimeridian=FALSE)

  df0$bathymetry <- get.depth(bathy, df0[,c("LON","LAT")], locator=F)$depth
```


```{r, eval=F}
# SAVE OCCURRENCE DATA
---------------------------------    

#NOTA INTERNA: quiero guardar el RData en la carpeta data/occurrences con el nombre de la especies sin espacios para no machar los RData en caso de que hagamos más de una especie pero no lo consigo. 

# lo dejo como esta, asi luego al cargar los datos no hay que poner el nombre de la especie y es mas sencillo

save(df0, file = "data/occurrences/occ.RData")
save(study_area, file = "data/spatial/total_area.RData")
```


```{r, eval=F}
# PLOT OCCURRENCES MAP
---------------------------------   
ggplot() +
   geom_path(data = study_area, 
             aes(x = long, y = lat, group = group),
             color = 'gray', size = .2) +
   geom_point(data=df0, aes(x=decimalLongitude, y=decimalLatitude,colour= occurrenceStatus)) 
  
```

## Create pseudo-absence data

Prevalence 50%

See code from ANICHO (mantaining some space around presences). Leire C.

Ref [@barbetmassin_etal_2012]

Copio aqui el codigo de anicho tal cual, luego lo limpiaré para este caso:

```{r, eval=F}
# Script information ------------------------------------------------------

# Title: Generate pseudo-absences for IM-18-ANICHO
# Last modified by Leire Ibaibarriaga (libaibarriaga@azti.es) and Leire Citores (lcitores@azti.es)

# Load libraries ----------------------------------------------------------

library(tidyverse)
library(ggplot2)
library(scales)
library(here)
library(ggridges)

library(maps)        # some basic country maps
library(mapdata)     # higher resolution maps
library(mapproj)
library(marmap)      # access global topography data
library(mapplots)    # ices rectangles
library(sf)
library(gridExtra)
library(lubridate)
library(raster)

# load data-files
#occurrence data
load("data/occurrences/occ.RData")
#map Atlantic without black sea
load("data/spatial/total_area.RData")

#remove points in land
df0<-subset(df0,bathymetry<0)

#use years from 2000 to 2014
df0<- subset(df0, YEAR<=2014 & YEAR>=2000)

# Remove points in south hemisphere (optional)
#prop.table(table(df0$LAT<=0)) #only 22% of obs in the south
#df0<-subset(df0,LAT>0)

# Remove south hemisphere area
#area <- crop(study_area, extent(-100, 100, 0, 100))
area <- study_area

#convert to spatial point data frame
df<-df0 ; coordinates(df)<- ~LON+LAT
crs(df)<-crs(area)

#convert to sf
area.sf<-st_union(st_as_sf(area))
df.sf<-st_as_sf(df)

ggplot(area.sf) + geom_sf() + geom_sf(data=st_union(df.sf))


dim(df) # 15079    5
head(df)
tail(df)
summary(df)
```


```{r, eval=F}
# Maps --------------------------------------------------------------------

# basic ggplot

global <- map_data("worldHires")

p0 <- ggplot() + 
  annotation_map(map=global, fill="grey")+
  geom_sf(data=area.sf,fill=NA)
print(p0)

# EGSP: Transformation to UTM ---------------------------------------------

# function to find your UTM. Taken from Nerea Goikoetxea

lonlat2UTM = function(lonlat) {
  utm = (floor((lonlat[1] + 180) / 6) %% 60) + 1
  if(lonlat[2] > 0) {
    utm + 32600
  } else{
    utm + 32700
  }
}

EPSG_2_UTM <- lonlat2UTM(c(mean(df$LON), mean(df$LAT))) 
EPSG_2_UTM # 32623

# transform to UTMs (in m)
aux <- st_transform(area.sf, EPSG_2_UTM)
df.sf.utm <- st_transform(df.sf, EPSG_2_UTM)

# number of points within the study area

df.in <- st_intersects(df.sf.utm, aux, sparse=FALSE)
df$INSIDE <- df.in[,1]
mean(df$INSIDE) # 100% points inside the area of study

# create buffers of 10km (10000) around the points and join the resulting polygons

buffer <- st_buffer(df.sf.utm, dist=100000)
buffer <- st_union(buffer)

p0 + geom_sf(data=buffer)

p <- p0 +
  geom_sf(data=buffer, fill="blue")+
  coord_sf(xlim=c(-90,-82), ylim=c(25,31))
print(p)

# intersect the result with the buffers around the catch data points
aux0 <- st_difference(aux, buffer)

# ggplot for all data

p <- p0 +
  geom_sf(data=aux0, fill=2)+
  coord_sf(xlim=c(-95,-82), ylim=c(25,31))
print(p)

p<-p0 + geom_sf(data=aux0,fill=2)

ggsave(file.path("plots","pseudo",paste0("area_pseudo_all.png")), p, device="png")
```


```{r, eval=F}
# Generate pseudo-absences -------------------------------------------------


# Generate the pseudo-absence data frame

pseudo <- matrix(data=NA, nrow=dim(df0)[1], ncol=dim(df0)[2])
pseudo <- data.frame(pseudo)
names(pseudo) <- names(df0)

# set the seed

set.seed(1)

##no loop

buffer.sub <- st_buffer(df.sf.utm, dist=100000)
buffer.sub <- st_union(buffer.sub)
aux0.sub <- st_difference(aux, buffer.sub)
rp.sf <- st_sample(aux0.sub, size=dim(df.sf.utm)[1], type="random") # randomly sample points
rp.sf <- st_transform(rp.sf, 4326)
rp <- as.data.frame(st_coordinates(rp.sf)) # transform to lat&lon and extract coordinates as data.frame
pseudo$LON <- rp$X
pseudo$LAT <- rp$Y
p <- p0 +
  geom_sf(data=aux0.sub, fill=2, alpha=0.3)+
  geom_sf(data=rp.sf, col=1, shape=4,size=0.5)+
  geom_sf(data=df.sf.utm, col=4, alpha=0.3)+
  ggtitle(paste(unique(df$scientificName),"all"))
ggsave(file.path("plots","pseudo",paste0("pseudo_","all",".png")), p, device="png")


p <- p0 +
  geom_sf(data=aux0.sub, fill=NA, alpha=0.3)+
  geom_sf(data=df.sf.utm, col=4, alpha=0.3,size=0.5)+
  geom_sf(data=rp.sf, col=1, shape=4,size=0.5)+
  coord_sf(xlim=c(-98,-79), ylim=c(15,31))+
  ggtitle(paste(unique(df$scientificName),"all"))
ggsave(file.path("plots","pseudo",paste0("pseudo_","all_zoom",".png")), p, device="png")


# complete the rest of columns

pseudo$scientificName <- df0$scientificName
pseudo$occurrenceStatus  <- 0
```


```{r, eval=F}
# Save the final dataset including the pseudo-absences --------------------

head(df0)
head(pseudo)

# Join the two data sets and save the final dataset

dat <- rbind(df0, pseudo)
save(list=c("dat"),file=file.path("data","outputs_for_modelling",file="PAdata.RData"))
save(list=c("area"),file=file.path("data","spatial",file="area.RData"))
write.table(dat, file=file.path("data","outputs_for_modelling","PAdata.csv"), row.names=F, sep=";", dec=".")

```



