---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Environmental data

In this chapter we first, download environmental data from a public repository; second, we crop the data to our area of interest, and we save it as raster stack. 

As in Chapter 2, first, we load a list of required libraries.
```{r, eval=T,message=FALSE,warning=FALSE}
requiredPackages <- c(
  #GENERAL USE LIBRARIES --------#
  "here", # Library for reproducible workflow
  "rstudioapi",  # Library for reproducible workflow
  "maptools", #plotting world map
  "ggplot2", #for plotting
  "knitr",  # format tables
  "kableExtra", # format tables
  "raster", # to work with spatial data
  "dplyr",
  
  #DOWNLOAD FROM PUBLIC REPOSITORIES --------#
  "sdmpredictors" #to access Bio-ORACLE dataset
  
    )
```

We run a function to install the required packages that are not in our system and load all the required packages.
```{r, eval=T,message=FALSE,warning=FALSE}
install_load_function <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

install_load_function(requiredPackages)
```

We define some overall settings.
```{r, eval=T,message=FALSE,warning=FALSE}
# General settings for ggplot (black-white background, larger base_size)
theme_set(theme_bw(base_size = 16))
```

## Download from public repositories

Environmental data can be available from different sources. In this case, we used the Bio-ORACLE (ocean
rasters for analyses of climate and environment) database [@tyberghein_etal_2012; @assis_etal_2017]. These data are publicly available and are easily accessible from the [`sdmpredictors` package](https://cran.r-project.org/web/packages/sdmpredictors/index.html). 

We can check the list of available datasets with the function `list_datasets` from `sdmpredictors` package, as follows: 
```{r, eval=T,message=FALSE,warning=FALSE}
mydat <- list_datasets()

kable(mydat)%>% 
  kable_styling("striped") %>% 
  scroll_box(height="600px", width = "100%")
```


By default this function returns all the supported datasets. To return only marine datasets we can set the `marine` argument equal to `TRUE` or equivalently we could set the `terrestrial` and `freshwater` arguments equal to `FALSE`:
```{r, eval=T,message=FALSE,warning=FALSE}
mydat <- list_datasets(marine=T)
# or equivalently: 
# mydat <- list_datasets(terrestrial=F, freshwater=F)
# mydat <- list_datasets(marine=T, terrestrial=F, freshwater=F)
```

There are two datasets (Bio-ORACLE and MARSPEC) that have marine data. The function `list_layers` returns information on the layers of one or more datasets. So, we can see the layers available in the Bio-ORACLE dataset as follows:
```{r, eval=T,message=FALSE,warning=FALSE}
mytab <- list_layers("Bio-ORACLE")

kable(mytab)%>% 
  kable_styling("striped") %>% 
  scroll_box(height="600px", width = "100%")
```


Once we identify the dataset and the layers we are interested on, we can extract their details from the list. In this case, we download data on chlorophyll, salinity, diffuse attenuation coefficient and sea surface temperature.
```{r, eval=T,message=FALSE,warning=FALSE}
target <- c("BO2_chlomean_ss",
            "BO2_salinitymean_ss",
            "BO_damean",
            "BO_sstmean")

# Extrat details from the list
myvars <- mytab %>% 
  dplyr::filter (mytab$layer_code %in% target)

myvars$name
```

And we can download them using the function `load_layers` 
```{r, eval=T,message=FALSE,warning=FALSE}
# Download layers
myBioracle.layers <- load_layers(c("BO2_chlomean_ss", "BO2_salinitymean_ss", "BO_damean" ,"BO_sstmean")) 

save (list = "myBioracle.layers", file = "data/env/myBioracle.layers.RData")
```

Note that the resulting object is a `rasterStack`, where each variable is a layer
```{r, eval=T,message=FALSE,warning=FALSE}
class(myBioracle.layers)
myBioracle.layers
``` 

And we can plot it:
```{r, eval=T,message=FALSE,warning=FALSE}
raster::plot(myBioracle.layers)
```

Note that the functions `dataset_citations` and `layer_citations` provide the bibliographic entries of the datasets and layers for proper citation:
```{r}
print(dataset_citations("Bio-ORACLE"))
print(layer_citations("BO2_chlomean_ss"))
```

In case we are not interested on the whole area, we can crop the raster objects to the area of interest. 

For example, we can load the `study_area` object that is a SpatialPolygonsDataFrame that has been created previously and defines the extent of our spatial data and we can crop the rasterStack to the same extent:
```{r, warning=F, message=F}
load(here::here ("data", "spatial", "study_area.RData"))

mylayers <- crop(myBioracle.layers, extent(study_area))

plot(mylayers)
```

To facilitate subsequent access, the rasterStack with the downloaded data is saved in a local folder: 
```{r, eval=T}
writeRaster(mylayers, filename="data/env/mylayers.tif", options="INTERLEAVE=BAND", overwrite=TRUE)
```





