---
output: html_document
editor_options: 
  chunk_output_type: console
---

<!-- 
This file is part of a gitbook that should be cited as: 

Valle, M., Citores, L., Ibaibarriaga, L., Chust, C. (2023) GAM-NICHE: Shape-Constrained GAMs to build Species Distribution Models under the ecological niche theory. AZTI. https://doi.org/10.57762/fzpy-6w51 

This tutorial has been supported by the European Union’s Horizon 2020 research and innovation programme under grant agreements No 862428 MISSION ATLANTIC project
-->

# Prepare the final dataset

In this chapter we first, extract environmental data associated to the presence/pseudo-absence data, we explore the data we got, we check correlation between variables and we calculate the Variance Inflation Factor (VIF) to make a selection of the variables we are going to use in the model. 

First, we load a list of required libraries.
```{r, eval=T,message=FALSE,warning=FALSE}
requiredPackages <- c(
  "here", 
  "rstudioapi",  
  "sp", 
  "raster", 
  "dplyr",
  "tidyr",
  "ggplot2",
  "ggcorrplot",
  "GGally", 
  "HH" 
    )
```

We run a function to install the required packages that are not in our system and load all the required packages.
```{r, eval=T,message=FALSE,warning=FALSE}
install_load_function <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

install_load_function(requiredPackages)

```

We define some overall settings.
```{r, eval=T,message=FALSE,warning=FALSE}
# General settings for ggplot (black-white background, larger base_size)
theme_set(theme_bw(base_size = 16))
```

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width=16, fig.height=12)
```

## Extract environmental data associated to species distribution data

Once we have prepared our species distribution data (occurrences and pseudo-absences) and the environmental rasters, we need to merge both sources of data. First, we load the objects created in previous sections:
```{r, eval=T,message=FALSE,warning=FALSE}
# Load presence-absence data
load(here::here ("data", "outputs_for_modelling", "PAdata.RData"))

# Load environmental rasters
mylayers<-stack(here::here ("data", "env", "mylayers.tif"))
```

Now we can extract the environmental data associated to each of the species data points using the function `extract` from the `raster` package. The method employed is `bilinear` that returns the interpolated value from the four nearest raster cells.
```{r, eval=T,message=FALSE,warning=FALSE}
raster_ex <- raster::extract(x=mylayers, y=PAdata[,c("LON","LAT")], method="bilinear", na.rm=TRUE, df=T) 

colnames(raster_ex)[-1]<-c("BO2_chlomean_ss", "BO2_salinitymean_ss", "BO_damean" ,"BO_sstmean")

head(raster_ex)
```

We merge the presence/pseudo-absence data and the environmental data: 
```{r, eval=T,message=FALSE,warning=FALSE}
data <- cbind(PAdata, raster_ex)
```

We can conduct some quick checks on the new dataset:  
```{r, eval=T,message=FALSE,warning=FALSE}
dim(data)
str(data)
head(data)
summary(data)
```

The new dataset has 29806 rows and 10 columns, and there are 139 NA's in the environmental dataset. We remove the points with NA's:
```{r, eval=T,message=FALSE,warning=FALSE}
# we remove year column because pseudoabsences miss this info
data <- data %>% 
  dplyr::select (-YEAR) %>% 
  na.omit()
```

We check again the dataset:
```{r, eval=T,message=FALSE,warning=FALSE}
dim(data)
summary(data) 
```

The resulting dataset has 29661. We save this dataset in a local file to work on it in subsequent steps. 
```{r, eval=T,message=FALSE,warning=FALSE}
save(list="data", file="data/outputs_for_modelling/PAdata_with_env.RData")
```

## Exploratory plots of environmental variables

Before starting the modelling process, we are going to explore the individual variables in the dataset. 

We can explore the distributions of each of the environmental variables by looking at the violin and boxplots and at the histograms and density plots as follows: 

```{r, eval=T,message=FALSE,warning=FALSE}
tmp <- data[, c("BO2_chlomean_ss","BO2_salinitymean_ss","BO_damean","BO_sstmean")]
tmp <- pivot_longer(data=tmp, cols=everything()) 

ggplot(data=tmp, aes(x=name, y=value)) + 
  geom_boxplot()+
  facet_wrap(~name, scales="free")

ggplot(data=tmp, aes(x=name, y=value)) + 
  geom_violin(fill="red", alpha=0.3)+
  geom_boxplot(width=0.1)+
  facet_wrap(~name, scales="free")

ggplot(data=tmp, aes(x=value)) + 
  geom_histogram(aes(y= after_stat(density)), colour=1, fill="red", alpha=0.3)+
  geom_density(lwd=1)+
  facet_wrap(~name, scales="free")
```

## Exploratory plots of environmental variables depending on species distribution data

To analyse if there are preferences for certain ranges of the environmental variables, we compare the distribution of the environmental variables for presence and pseudo-absence data:   
```{r, eval=T,message=FALSE,warning=FALSE}
tmp <- data[, c("LON", "LAT", "BO2_chlomean_ss","BO2_salinitymean_ss","BO_damean","BO_sstmean","occurrenceStatus")]
tmp <- pivot_longer(data=tmp, cols=!occurrenceStatus) 

ggplot(data=tmp, aes(x=factor(occurrenceStatus), y=value, fill=factor(occurrenceStatus), group=factor(occurrenceStatus))) + 
  geom_violin(alpha=0.3)+
  geom_boxplot(fill="white", width=0.1)+
  facet_wrap(~name, scales="free")+
  theme(legend.position = "bottom",legend.background = element_rect(fill = "white", colour = NA))

ggplot(data=tmp, aes(x=value, fill=factor(occurrenceStatus), group=factor(occurrenceStatus))) + 
  geom_density(lwd=1, alpha=0.3)+
  facet_wrap(~name, scales="free")+
  theme(legend.position = "bottom",legend.background = element_rect(fill = "white", colour = NA))
```

## Correlation analysis

Some of the environmental variables can be correlated. The `GGally` package allows to easily produce pairplots of the variables and their correlation.  
```{r, eval=T,message=FALSE,warning=FALSE}
tmp <- data[, c("LON","LAT","BO2_chlomean_ss","BO2_salinitymean_ss","BO_damean","BO_sstmean")]

ggpairs(tmp) #this takes some minutes
```

A more detailed analysis of the potential correlations can be conducted using the package `ggcorrplot`:
```{r, eval=T,message=FALSE,warning=FALSE}
mat <- cor(tmp, use="complete.obs") 
p.mat <- cor_pmat(tmp)

ggcorrplot(mat, type = "lower", lab=T, p.mat = p.mat)
```

## Variance Inflation Factor (VIF) 

Furthermore, multicollinearity in regression analysis can be explored using the VIF (Variance Inflation Factor). The value of the VIF statistics indicate the level of multicollinearity with the rest of the variables:

* VIF equal to 1 = variables are not correlated
* VIF between 1 and 5 = variables are moderately correlated 
* VIF greater than 5 = variables are highly correlated

There are several packages in `R` that allows to calculate the VIF statistics. In this case we use the package `HH`: 
```{r, eval=T,message=FALSE,warning=FALSE}
# Select variables for VIF calculation
v.table <- data %>% 
  dplyr::select (BO2_salinitymean_ss, BO_sstmean, BO2_chlomean_ss, BO_damean)

# Get VIF results
out.vif <- vif(v.table)
sort(out.vif)
```

We remove the variable that has the highest VIF value and we test again the multicollinearity: 
```{r, eval=T,message=FALSE,warning=FALSE}
v.table <- v.table %>% 
  dplyr::select (-BO_damean)

# Get new VIF results
out.vif <- vif(v.table)
sort(out.vif)
```

Now all the variables have VIF values that are acceptable. So, we proceed to remove BO_damean (Diffuse attenuation coefficient at 490 nm). And save the selected variables for the next modelling stages:
```{r, eval=T,message=FALSE,warning=FALSE}
data <- data %>% dplyr::select (-BO_damean)
```

We save the dataset as our output for modelling.
```{r, eval=T,message=FALSE,warning=FALSE}
save(list="data", file="data/outputs_for_modelling/PAdata_with_env.RData")
```


