---
output: html_document
editor_options: 
  chunk_output_type: console
---

<!-- 
This file is part of a gitbook that should be cited as: 

Valle, M., Citores, L., Ibaibarriaga, L., Chust, C. (2023) GAM-NICHE: Shape-Constrained GAMs to build Species Distribution Models under the ecological niche theory. AZTI. https://doi.org/10.57762/fzpy-6w51 

This tutorial has been supported by the European Union’s Horizon 2020 research and innovation programme under grant agreements No 862428 MISSION ATLANTIC project
-->

# Shape Constrained-Generalized Additive Models

In order to fit SDM in agreement with the ecological niche theory, the proposed Shape-Constrained Generalized Additive Models (SC-GAMs) in [@citores_etal_2020] are fitted in this section. SC-GAMs are based on Generalized Additive Models, allowing us to impose shape-constraints to the linear predictor function. The R package SCAM implements the general framework developed by [@pya_etal_2015] using shape-constrained P-splines. Monotonicity and concavity/convexity constraints can be imposed on the sign of the first and/or the second derivatives of the smooth terms. For fitting Species Distribution Models in agreement with the ecological niche theory, we imposed concavity  constraints ($f''(x) \le 0$), so that the response can presents at most a single mode.  

Alternatively, the R package `mboost` fits SC-GAMs using boosting methods. We are not going to develop this alternative here.

First, we load the list of required libraries.
```{r, eval=T,message=FALSE,warning=FALSE}
requiredPackages <- c(
  #GENERAL USE LIBRARIES --------#
  "here", # Library for reproducible workflow
  "rstudioapi",  # Library for reproducible workflow
  "stringr",
  "RColorBrewer",  
  "ggplot2",
  "dplyr",
  
  #SPATIAL DATA --------#
  #"rgdal",
  "fields",
  "maps" ,
  "raster",

  #MODEL FIT --------#
  "scam",
  "plotmo",
  "pkgbuild",
  "dismo"
  )
```

We run a function to install the required packages that are not in our system and load all the required packages.
```{r, eval=T,message=FALSE,warning=FALSE}
install_load_function <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

install_load_function(requiredPackages)

```

We define some overall settings.
```{r, eval=T,message=FALSE,warning=FALSE}
# General settings for ggplot (black-white background, larger base_size)
theme_set(theme_bw(base_size = 16))
```

## Model fit

We set the working directory to the folder where the current script is located and we load the dataset (PAdata_with_env.Rdata) containing the presence-absence data together with the environmental data.
```{r, eval=T,message=FALSE,warning=FALSE}
load(here::here ("data", "outputs_for_modelling", "PAdata_with_env.Rdata"))
```

To fit a logistic regression model in the SC-GAMs framework, we use the scam function, where we set the binomial family with the logit link function. Our response variable is the presence-absence data and the selected three explanatory variables are the SST, chlorophyll and salinity. Each variable is included in the model through an spline function where the concavity constraint is set using bs="cv". The details about this option can be found in the section "Constructor for concave P-splines in SCAMs" of the SCAM manual (https://cran.r-project.org/web/packages/scam/scam.pdf). The number of knots (k) is fixed at 8 in this example for a good balance between flexibility and computation time.

**UNIVARIATE MODELS**

Before fitting the model with the selected three environmental variables, we can fit univariate model as follows.

We fit the univariate model for SST, we print the summary of the model fit, and look at the fitted curve in the response scale.
```{r, eval=T,message=FALSE,warning=FALSE}
model_sst <- scam (occurrenceStatus ~  s(BO_sstmean, k=8,bs="cv"), family=binomial(link="logit"), data=data)
summary(model_sst)
plotmo(model_sst,level = 0.95, pt.col=8)
```

We repeat the same steps for the rest of the variables.
```{r, eval=T,message=FALSE,warning=FALSE}
model_chl <- scam (occurrenceStatus ~  s(BO2_chlomean_ss, k=8,bs="cv"), family=binomial(link="logit"), data=data)
summary(model_chl)
plotmo(model_chl,level = 0.95, pt.col=8)
```

Due to convergence issues, sometimes it is necessary to fix the smoothing parameter (sp) at small value, i.e. $10^{-5}$, as here when introducing salinity as an explanatory variable. If no value is provided, the smoothing parameter is estimated within the model.
```{r, eval=T,message=FALSE,warning=FALSE}
model_sal <- scam (occurrenceStatus ~  s(BO2_salinitymean_ss, k=8,bs="cv"), family=binomial(link="logit"), data=data,sp=0.00001)
summary(model_sal)
plotmo(model_sal,level = 0.95, pt.col=8)
```

**MODEL WITH ALL VARIABLES**

Now we fit the model including the three selected variables.
```{r, eval=T,message=FALSE,warning=FALSE}
model <- scam (occurrenceStatus ~  s(BO_sstmean, k=8,bs="cv")+ s(BO2_salinitymean_ss, k=8,bs="cv")+s(BO2_chlomean_ss, k=8,bs="cv"), family=binomial(link="logit"), data=data,sp=rep(0.00001,3))
summary(model)
plotmo(model,level = 0.95, pt.col=8)
```

We can see in the summary of the fit, that all included variables are statistically significant (with p<0.05) and present a unimodal response curve.

For a more detailed check of the fitting, we can use the scam.check function:
```{r, eval=T,message=FALSE,warning=FALSE}
old.par <- par(mfrow=c(2,2))
scam.check(model)
```

## Model selection

When several explanatory variables are available, a variable selection process can be carried out. Here we provide as an example, a function that performs forward variable selection (modsel.scam) based on the significance of variables and AIC values of the fits. 
```{r, eval=T,message=FALSE,warning=FALSE}
source("function/function_scam_selection_optimized.R")
```

We save the names of the variables we want to introduce for the variable selection process as a vector:
```{r, eval=T,message=FALSE,warning=FALSE}
vars <- c("BO_sstmean",
          "BO2_salinitymean_ss",
          "BO2_chlomean_ss")
```

The default AIC tolerance is 2 and there is not a limit on selected terms in this example. These options can be modified through aic.tol and vmax arguments in the function. The number of knots and the sp can be also modified.
```{r, eval=T, message=FALSE,warning=FALSE}
model_SCGAM <- try(modsel.scam(basef="occurrenceStatus ~ 1", vars=vars, dat=data,sp=0.00001), silent=T)  
```

We check results of the selected model, such as, selected variable names:
```{r, eval=T, message=FALSE,warning=FALSE}
model_SCGAM$svars
```

AICs of the fitted models:
```{r, eval=T, message=FALSE,warning=FALSE}
sapply(model_SCGAM$smod, AIC)
```

Explained deviance of fitted models:
```{r, eval=T, message=FALSE,warning=FALSE}
sapply(model_SCGAM$smod, function(x) summary(x)$dev.expl)
```

Formulas of the fitted models:
```{r, eval=T,message=FALSE,warning=FALSE}
lapply(model_SCGAM$smod, formula)
```

Summaries of the fitted models:
```{r, eval=T,message=FALSE,warning=FALSE}
lapply(model_SCGAM$smod, summary)
```

The last model of the list, is the selected one. Which in this case contains the considered three variables.
```{r, eval=T,message=FALSE,warning=FALSE}
selected_model <- model_SCGAM$smod[[length(model_SCGAM$smod)]]
```

We save the selected model
```{r, eval=T}
save(list="selected_model", file="models/selected_model.RData")
```

Note that there are multiple options and criteria for model selection that are not reviewed here. Any model selection technique used for GAMs can be used also for SC-GAMs.