---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Shape Constrained-Generalized Additive Models

One citation is [@citores_etal_2020]

Mention there is an alternative using `mboost` that won't be further developed here.  

Loading required libraries

```{r, eval=F}
library(scam)
#library(mgcv)
library(plotmo)
library(rgdal)
library(ggplot2)
library(dplyr)
library(fields) #imageplot
library(maps) #map world
library(raster)
library(RColorBrewer)#color palette

#SDMTools - install RTools and follow: https://cran.r-project.org/bin/windows/Rtools/
#install.packages("remotes")
#remotes::install_version("SDMTools", "1.1-221.2")
library(SDMTools)
library(dismo)
library(stringr)
library(rstudioapi)
```


## Model fit


```{r, eval=F}

# location of script
setwd(dirname(getSourceEditorContext()$path))

#NOTA INTERNA: aquí hay que cargar el RData de presencias y pseudoausencias que creemos en 05_data_final de momento he copido los datos de PA de la anchoa para poder correr el código. Habrá que cambiar status por el nomnbre de la columna de presencia-pseudoausencia y el nombre de las variables, en este caso lo hacemos para la anchoa con todas las variables que tenemos: 

#Load presence-absence with env data dataset
load(file.path ("data", "outputs_for_modelling", "PAdata_with_env.Rdata"))

#UNIVARIATE MODELS
#sst
model0 <- scam (occurrenceStatus ~  s(BO_sstmean, k=8,bs="cv"), family=binomial(link="logit"), data=data)
summary(model0)
plotmo(model0,level = 0.95, pt.col=8)

#chlorophyll
model0 <- scam (occurrenceStatus ~  s(BO2_chlomean_ss, k=8,bs="cv"), family=binomial(link="logit"), data=data)
summary(model0)
plotmo(model0,level = 0.95, pt.col=8)

model0 <- scam (occurrenceStatus ~  s(BO_bathymean, k=8,bs="cv"), family=binomial(link="logit"), data=data)
summary(model0)
plotmo(model0,level = 0.95, pt.col=8)

model0 <- scam (occurrenceStatus ~  s(BO2_salinitymean_ss, k=8,bs="cv"), family=binomial(link="logit"), data=data,sp=0.00001)
summary(model0)
plotmo(model0,level = 0.95, pt.col=8)

# including all variables
model <- scam (occurrenceStatus ~  s(BO_sstmean, k=8,bs="cv")+ s(BO2_salinitymean_ss, k=8,bs="cv")+s(BO2_chlomean_ss, k=8,bs="cv")+s(BO_bathymean,k=8,bs="cv"), family=binomial(link="logit"), data=data,sp=rep(0.00001,4))

# including all variables except bathymetry
model_nobathy <- scam (occurrenceStatus ~  s(BO_sstmean, k=8,bs="cv")+ s(BO2_salinitymean_ss, k=8,bs="cv")+s(BO2_chlomean_ss, k=8,bs="cv"), family=binomial(link="logit"), data=data,sp=rep(0.00001,3))


#Check model which contains all variables
summary(model)
scam.check(model)
plot(model)
plotmo(model,level = 0.95, pt.col=8)

summary(model_nobathy)
plotmo(model_nobathy,level = 0.95, pt.col=8)

# Save the model
selected_model<-model
save(list=c("model","model_nobathy","selected_model"), file = ("models/model.RData"))
```


## Model selection

Source the function for forward selection of the variables developed by libaibarriaga criterium: AIC and pvalue

```{r, eval=F}
source("function/function_scam_selection_optimized.R")
```

preparing data
```{r, eval=F}
# selecting the variables to be used in the model 

#NOTA INTERNA: añadir el nombre de las variables que tenemos
vars <- c("BO_sstmean",
          "BO2_salinitymean_ss",
          "BO2_chlomean_ss",
          "BO_bathymean")
```

Code to build SC-GAM model default sp= NULL, k = 8 and plevel  = 0.05
```{r, eval=F}
  # apply the model selection function (no limits in the number of terms)
  model_SCGAM <- try(modsel.scam(basef="occurrenceStatus ~ 1", vars=vars, dat=data,sp=0.00001), silent=T) # by default aic.tol=2
#NOTA INTERNA: ¿queremos añadir el if else y asignar un valor de sp en caso de que sp NULL de error?

```

check results
```{r, eval=F}
model_SCGAM$svars
sapply(model_SCGAM$smod, AIC)
sapply(model_SCGAM$smod, function(x) summary(x)$dev.expl)
lapply(model_SCGAM$smod, formula)
lapply(model_SCGAM$smod, summary)

#Select the last model of the list which is the best model according to AIC and plevel criteria
selected_model <- model_SCGAM$smod[[length(model_SCGAM$smod)]]
selected_model$sp

# see specifically the summary and the plots of the final model after the selection
# note that it is the last element of the list of models
summary(selected_model)
scam.check(selected_model)
plot(selected_model)
plotmo(selected_model,level = 0.95, pt.col=8)



# Save the model

#save(selected_model, file = "models/selected_model.RData")

```
